# Quantum channels (or CP maps) {#chapter10}

> About linear transformations --- often called **quantum channels**, **quantum operations**, or **superoperators** --- that map density operators to density operators. Mathematically, such objects are known as **completely positive trace-preserving maps**. Be prepared for some name dropping; you will hear about Karl Kraus, Woody Stinespring, Andrzej Jamiołkowski, and Man-Duen Choi. To be sure, knowing names will not give you more insights, but at least you will not be intimidated when you hear about the **Stinespring** and the **Kraus representations**, the **Jamiołkowski isomorphism**, or the **Choi matrix**.

In quantum information science (and in life) we often divide the world into two parts: things we can control and everything else.
Things we can control (at least with some precision), such as a bunch of ions in an ion trap, will now acquire a generic name -- the **system** -- and the "everything else" part will be called the **environment**.
The joint evolution of the system and the environment is unitary, but the system _alone_ may deviate from the unitary evolution due to the entangling interactions with the environment.
What does this non-unitary quantum evolution look like?

Mathematically speaking^[Recall that, given a pair of Hilbert spaces $\mathcal{H}_A$ and $\mathcal{H}_B$, we denote the set of (bounded) linear operators from $\mathcal{H}_A$ to $\mathcal{H}_B$ by $\mathcal{B}(\mathcal{H}_A,\mathcal{H}_B)$. The shorthand $\mathcal{B}(\mathcal{H})$ is used to denote $\mathcal{B(H,H)}$. For example, an element of $\mathcal{B}(\mathcal{H})$ is exactly a density operator on $\mathcal{H}$.], we want to characterise **physically admissible** maps
$$
  \mathcal{B}\colon \mathcal{B}(\mathcal{H}) \to \mathcal{B}(\mathcal{H}')
$$
that map density operators to density operators.
We will call them **quantum channels**.
Here, $\mathcal{H}$ and $\mathcal{H}'$ are the Hilbert spaces associated with the input and the output, respectively (these two spaces may be of different dimensions: $\dim\mathcal{H}=d$ and $\dim\mathcal{H}'=d'$, with $d$ not necessarily equal to $d'$).
It is not a trivial task, since some operations (such as the matrix transpose) turn density matrices into seemingly legal density matrices, and yet they are _not_ physically admissible.
We want to understand why.

There are essentially two approaches to characterising quantum channels. Let us call them **constructive** and **axiomatic**.

- In the constructive approach (much favoured by physicists) we assume that, in the quantum world, there are only unitary evolutions, and that _non_-unitary evolutions of subsystems are induced by unitary evolutions of _larger_ systems.
We can view them as the result of adding or removing physical systems that participate in unitary evolutions.
In particular, _any_ quantum channel can be constructed by taking a physical system in state $\rho$, combining it with an auxiliary system (e.g. the environment) in some fixed quantum state, applying a unitary transformation to the combined system, and then discarding everything that is of no interest to us.

- In the axiomatic approach we proceed as mathematicians usually do.
We refer to quantum theory for inspiration, scratch our heads, and list all the desired properties that a map should have in order to be called an admissible quantum operation.
This way we define **completely positive trace-preserving** (CPTP) maps.

In the end, you will be relieved to learn, the two approaches are equivalent, leading to the same mathematical formalism.
This equivalence is often goes by the name of the **Stinespring dilation theorem**^[[William Forrest "Woody" Stinespring](https://en.wikipedia.org/wiki/W._Forrest_Stinespring) (1929--2012) was an American mathematician specialising in operator theory.], which says that the CPTP maps are exactly the maps which can be constructed from unitary evolutions by adding and removing subsystems.






## Unitary evolution is all there is...

Here we consider the constructive approach described above.
So we know that any evolution of any _isolated_ system is always unitary, since, at the fundamental level, this is the only evolution that is offered by quantum theory.
Anything else that is not unitary has to be, somehow, derived from a unitary evolution of the whole, and it is easy to see that a unitary evolution of the whole does not imply unitary evolutions of its constituent parts.
The question is: _what kind of sub-evolutions can be induced on subsystems by a global unitary evolution?_

Consider a physical system, prepared in some input state $\ket{\psi}\in\mathcal{H}$, which is then combined with another system (here, the environment) that is initially in state $\ket{e}\in\mathcal{H}_E$.
The two systems interact, and their joint unitary evolution $U$ produces, in general, an entangled state.
The system itself, after discarding the environment, ends in an output state $\rho'$, which is given by taking a partial trace, as shown in Figure \@ref(fig:joint-system-discard-output).

(ref:joint-system-discard-output-caption) The output (discarding what happens to the environment) is given by $\rho'=\tr_E \big(U\proj{\psi}\otimes\proj{e} U^\dagger\big)$.

```{r joint-system-discard-output,engine='tikz',engine.opts=list(template="tikz2pdf.tex"),fig.width=6,fig.cap='(ref:joint-system-discard-output-caption)'}
\begin{equation*}
  \begin{quantikz}
    \lstick{input $\ket{\psi}$} & \gate[wires=2]{U} &\rstick{output $\rho'$} \qw
  \\\lstick{fixed $\ket{e}$} &\hphantom{really wide} &\rstick{discard} \qw
  \end{quantikz}
\end{equation*}
```

The requirement that the environment start in a pure state is not very restrictive: if it were initially in a mixed state, then we could always regard the environment as a sub-environment of some larger environment in an entangled pure state.
It _is_ important, however, that the system and its environment _initially have no correlation with each other_: they are in a product state $\ket{\psi}\otimes \ket{e}$.
Then, if we fix the initial state of the environment $\ket{e}$ and the unitary $U$, the final state of the system depends _only_ on its initial state $\ket{\psi}$.
In fact, the initial state of the system can be a mixed state, described by some density operator $\rho$.
This is because any mixed state can be interpreted as a statistical mixture of pure states, and any linear process respects such a mixture.
Thus, by fixing $U$ and $\ket{e}$, we have constructed a well defined linear map $\mathcal{E}\colon\mathcal{B}(\mathcal{H})\to\mathcal{B}(\mathcal{H})$ that transforms the input state $\rho$ into the output state $\rho'=\mathcal{E}(\rho)$.

::: {.idea data-latex=""}
$$
  \begin{aligned}
    \mathcal{E}\colon \mathcal{B}(\mathcal{H}) &\longrightarrow \mathcal{B}(\mathcal{H})
  \\\rho &\longmapsto \rho'=\tr_E \big(U\proj{\psi}\otimes\proj{e} U^\dagger\big)
  \end{aligned}
$$
:::

This way of describing the action of a quantum channel is often called the **unitary representation** of a channel.
As mentioned above, the channel is completely defined by the unitary $U$ and the initial state of the environment $\ket{e}$, but note that, once the system and the environment cease to interact, any operation _on the environment alone_ has no effect on the state of the system.
That is, the two diagrams in Figure \@ref(fig:environment-same-channel) define the same channel, or, in other words, the unitaries $U$ and $(\id\otimes R)U$, where $R$ acts only on $\mathcal{H}_E$, define the same channel.^[Consider $U$ and $U(\id\otimes R)$, where $R$ acts only on $\mathcal{H}_E$. Do these two unitaries also define the same quantum channel?]

(ref:environment-same-channel-caption) The $\bullet$ here represents discarding the final state of the environment.

```{r environment-same-channel,engine='tikz',engine.opts=list(template="tikz2pdf.tex"),fig.width=6,fig.cap='(ref:environment-same-channel-caption)'}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{\psi}$} & \gate[wires=2]{U} &\qw &\qw &\rstick{$\rho'$} \qw
  \\\lstick{$\ket{e}$} &\hphantom{really wide} &\qw &\qw &\rstick{$\bullet$} \qw
  \end{quantikz}
\end{equation*}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{\psi}$} & \gate[wires=2]{U} &\qw &\rstick{$\rho'$} \qw
  \\\lstick{$\ket{e}$} &\hphantom{really wide} &\gate{R} &\rstick{$\bullet$} \qw
  \end{quantikz}
\end{equation*}
```

Note that, for any operator $\tilde{\rho}$ on $\mathcal{H}\otimes\mathcal{H}_E$, and for any operator $R$ on $\mathcal{H}_E$, we have that
$$
  \tr_E\left[
    (\id\otimes R) \tilde{\rho} (\id\otimes R^\dagger)
  \right] = \tr_E \tilde{\rho}.
$$
The reduced density operator $\rho=\tr_E\tilde{\rho}$ is not affected by $R$.
We can easily prove this for operators $\tilde{\rho}$ that are tensor products $\tilde{\rho}=X\otimes Y$ (a good exercise to try yourself), and then, by linearity, extend the result to any operator $\tilde{\rho}$.

We must point out that our construction of quantum channels does not capture all possible quantum evolutions.
The assumption that the system and the environment are not initially correlated is crucial, and it does impose some restrictions on the applicability of our formalism.
Compare, for example, the following two versions of a process in which the system, initially in state $\rho$, undergoes two stages of evolution.

Firstly:

```{r,engine='tikz',engine.opts=list(template="tikz2pdf.tex"),fig.width=7}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\rho$} &\gate[2]{U_A} &\rstick{$\rho'$} \qw & &\gate[2]{U_B} &\rstick{$\rho''$} \qw
  \\\lstick{$\ket{e}$} &\qw &\qw &\qw &\qw & \ar[d,dash,thickness]\qw
  \\& & & & &\bullet
  \end{quantikz}
\end{equation*}
```

Here, the environment is not discarded after the first unitary evolution $U_A$ --- it carries on and participates in the second unitary evolution $U_B$.
In this case, the evolutions $\rho\mapsto\rho'$ and $\rho\mapsto\rho''$ _are_ well-defined quantum channels, but the evolution $\rho'\mapsto\rho''$ is _not_; it falls outside the remit of our formalism for the input state of the system and the state of the environment are not independent.

Secondly:

```{r,engine='tikz',engine.opts=list(template="tikz2pdf.tex"),fig.width=7}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\rho$} &\gate[2]{U_A} &\qw &\rstick{$\rho'$} \qw & &\qw &\gate[2]{U_B} &\rstick{$\rho''$} \qw
  \\\lstick{$\ket{a}$} & &\qw \ar[d,dash,thickness]\qw & & & \lstick{$\ket{b}$} &\qw &\ar[d,dash,thickness]\qw
  \\& &\bullet & & & & &\bullet
  \end{quantikz}
\end{equation*}
```

In contrast to the previous process, here we discard the environment after the first unitary, and start the second unitary evolution in a fresh tensor product state with a _new_ environment --- the two stages involve independent environments.
In this case, all three evolutions ($\rho\mapsto\rho'$, $\rho'\mapsto\rho''$, and $\rho\mapsto\rho''$) are well-defined quantum channels, and can be composed.^[Unitary evolutions form a group; quantum channels form a _[semigroup](https://en.wikipedia.org/wiki/Semigroup)_. Quantum channels are invertible only if they are unitary operations or simple isometric embeddings, such as including the environment in some fixed state and then discarding it right away, without any intermediate interaction.]
If $\mathcal{E}_A$ describes the evolution from $\rho$ to $\rho'$, and $\mathcal{E}_B$ the evolution from $\rho'$ to $\rho''$, then the composition $\mathcal{E}_B\mathcal{E}_A$ describes the evolution from $\rho$ to $\rho''$.

In practice, we often deal with complex environments that have internal dynamics that "hide" any entanglement with the system as quickly as it arises.
For example, let our system be an atom, surrounded by electromagnetic field (which serves as the environment), and let the field start in the vacuum state.
If the atom emits a photon into the environment, then the photon quickly propagates away, and the immediate vicinity of the atom appears to be empty, i.e. it resets to the vacuum state.
In this approximation, we assume that the environment quickly forgets about its state resulting from any previous evolution.
This is known as the **Markov approximation**, and, in the quantum Markov process^[[Andrey Markov](https://en.wikipedia.org/wiki/Andrey_Markov) (1929--2012) was a Russian mathematician best known for his work on stochastic processes.], the environment has essentially no memory.






## Operator sum representation

The unitary representation of a quantum channel is usually too general to be of any practical use, at least for our purposes.
We will find it convenient, as you will see when we start discussing quantum error correction, to choose an orthonormal basis $\{\ket{e_k}\}$ of $\mathcal{H}_E$, and express the joint (i.e. "system plus environment") unitary evolution $U$ as
$$
  U(\ket{\psi}\otimes\ket{e})
  = \sum_k E_k\ket{\psi}\otimes\ket{e_k}
$$
where $E_k$ is an operator on $\mathcal{H}$ defined by
$$
  E_k\ket{\psi}
  = \bra{e_k}U(\ket{\psi}\otimes\ket{e})
$$
for any vector $\ket{\psi}\in\mathcal{H}$
Think about $\bra{e_k}U(\ket{\psi}\otimes\ket{e}$ as a partial inner product^[Recall that, given $\mathcal{H}_A\otimes\mathcal{H}_B$, the partial inner product of $\ket{x}\in \mathcal{H}_A$ and $\ket{a}\otimes\ket{b}\in\mathcal{H}_A\otimes\mathcal{H}_B$ is first defined by the formula $$\bra{x}(\ket{a}\otimes\ket{b}) = \braket{x|a}\ket{b}$$ and then extended to other (non direct tensor product) vectors in  $\mathcal{H}_A\otimes \mathcal{H}_B$ by linearity.] of $\ket{e_k}\in\mathcal{H}_E$ and  $U\ket{\psi}\ket{e}\in \mathcal{H}\otimes\mathcal{H}_E$.
In analogy with classical communication channels we sometimes refer to the operators $E_k$ as **quantum errors**.

The intuition behind the expression
$$
  \ket{\psi}\otimes \ket{e}
  \longmapsto \sum_k E_k\ket{\psi}\otimes\ket{e_k}
$$
is that the operation $E_k$ (which changes the state of the system as $\ket{\psi}\mapsto E_k\ket{\psi}$) occurs with the probability $p_k=\bra{\psi}E^\dagger_k E_k\ket{\psi}$.
The probabilities sum up to one, i.e. $\sum_k p_k=\bra{\psi}E^\dagger_k E_k\ket{\psi} =1$ for any state $\ket{\psi}$, and so the operators $E_k$ must satisfy the completeness relation
$$
  \sum_k E^\dagger_k E_k=\id.
$$
