# (PART) Applications {-}

# Quantum algorithms {#quantum-algorithms}

<div class="video" title="Overview" data-videoid="pzC4pUK8s8A"></div>

> About quantum interference in disguise: Hadamard, **function evaluation**, Hadamard.
> Also about the early quantum algorithms and how they deal with querying oracles, searching for a needle in a haystack, and estimating periodicity of certain functions.
> Finally, about phase estimation, hidden order determination, and Shor's famous algorithm for prime factorisation, via the (inverse) quantum Fourier transform.

To boil down the theory of classical computers to a single sentence, we can say that they essentially evaluate functions: given $n$-bits of input, they produce $m$-bits of output that are uniquely determined by the input.
In other words, (very simple) classical computers encode binary functions
$$
  f\colon \{0,1\}^n \to \{0,1\}^m
$$
and then compute the value of the output for any particular specified $n$-bit argument.
But we can make an even further simplification: a binary function with an $m$-bit output value is equivalent to $m$-many binary functions with $1$-bit output values (which we call **Boolean functions**).
In other words, we might just as well say that the basic task performed by a computer is the evaluation of Boolean functions
$$
  f\colon \{0,1\}^n \to  \{0,1\}.
$$
How can we adapt this to the world of quantum computing?


## Quantum Boolean function evaluation

<div class="video" title="Abstract computation and reversible computation" data-videoid="8ok34yT-G4g"></div>

<div class="video" title="Quantum function evaluation" data-videoid="x_umQRMcwtA"></div>

In quantum computation, *all elementary operations are reversible* (i.e. unitary), so we need to compute Boolean functions in a reversible fashion --- we can do so as follows:
$$
  \ket{x}\ket{y} \longmapsto \ket{x}\ket{y\oplus f(x)}.
$$

The corresponding circuit diagram (for an input register of $n=3$ qubits) is shown in Figure \@ref(fig:n-equals-3-circuit-diagram).

(ref:n-equals-3-circuit-diagram-caption) Computing some $f\colon\{0,1\}^3\to\{0,1\}$ in a quantum manner, where $x\in\{0,1\}^3$, $y\in\{0,1\}$, and $\oplus$ denotes $\texttt{XOR}$, or addition modulo $2$.

```{r n-equals-3-circuit-diagram,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex"),fig.cap='(ref:n-equals-3-circuit-diagram-caption)'}
\begin{equation*}
  \begin{quantikz}
    \lstick[wires=3]{$\ket{x}$}
    & \ctrl{3}
    & \qw \rstick[wires=3]{$\ket{x}$}
  \\[-1em]
    & \ctrl{2}
    & \qw
  \\[-1em]
    & \ctrl{1}
    & \qw
  \\\lstick{$\ket{y}$}
    & \gate{f}
    & \qw \rstick{$\ket{y\oplus f(x)}$}
  \end{quantikz}
\end{equation*}
```

Here we use two registers: the first^[Reading the circuit diagram from top to bottom.] one stores the arguments $\ket{x}$ (where $x\in\{0,1\}^n$ is our binary string input), and the second one the value $f(x)$.
More precisely, the value $f(x)$ is added bit-wise to the pre-existing binary value $y$ of the second register.
We usually set $y=0$ to get
$$
  \ket{x}\ket{0} \longmapsto \ket{x}\ket{f(x)}.
$$

Quantum Boolean function evaluation is a special case of the generalised $x$-controlled-$U$ on two registers:
$$
  \sum_{x\in\{0,1\}^n} \proj{x}\otimes U_x
$$
where $U_x$ is either the identity $\id$ (when $f(x)=0$) or the bit-flip^[Do not confuse the capital $X$ (the Pauli bit-flip operator $\sigma_x$) with the small $x$ (a binary string stored in the first register, and the argument of our Boolean function $f$).] $X$ (when $f(x)=1$).
We can write this very succinctly as
$$
  \sum_{x\in\{0,1\}^n} \proj{x}\otimes X^{f(x)}.
$$
Because of this, we sometimes denote the quantum evaluation of the function $f$ by $U_f$, which is a gate on the $(n+1)$ qubits $\ket{x}\ket{y}$.


### A worked example

Consider the Boolean function $f\colon\{0,1\}^2\to\{0,1\}$ given by
$$
  f(x)
  = \begin{cases}
    1 &\text{if $x=01$;}
  \\0 &\text{otherwise}
  \end{cases}
$$
which we might call the **indicator** (or **characteristic**) function for the binary string $01$, sometimes denoted $\chi_{01}$.
The evaluation $\ket{x}\ket{y} \mapsto \ket{x}\ket{y\oplus f(x)}$ can be tabulated explicitly:
$$
  \begin{array}{cc}
    \ket{00}\ket{0} \longmapsto \ket{00}\ket{0}
    & \ket{00}\ket{1} \longmapsto \ket{00}\ket{1}
  \\\ket{01}\ket{0} \longmapsto \ket{01}\ket{1}
    & \ket{01}\ket{1} \longmapsto \ket{01}\ket{0}
  \\\ket{10}\ket{0} \longmapsto \ket{10}\ket{0}
    & \ket{10}\ket{1} \longmapsto \ket{10}\ket{1}
  \\\ket{11}\ket{0} \longmapsto \ket{11}\ket{0}
    & \ket{11}\ket{1} \longmapsto \ket{11}\ket{1}
  \end{array}
$$
and then
$$
  \begin{aligned}
    \sum_{x\in\{0,1\}^2} \ket{x}\bra{x}\otimes X^{f(x)}
    = &\proj{00} \otimes \id
    + \proj{01} \otimes X
  \\+ &\proj{10} \otimes \id
    + \proj{11} \otimes \id.
  \end{aligned}
$$

Finally, the matrix form looks as follows:
$$
  \left[
  \,
    \begin{array}{c|c|c|c}
      \mqty{1&0\\0&1}
      & \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
    \\\hline
    \mqty{0&0\\0&0}
      & \mqty{0&1\\1&0}
      & \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
    \\\hline
    \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
      & \mqty{1&0\\0&1}
      & \mqty{0&0\\0&0}
    \\\hline
    \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
      & \mqty{0&0\\0&0}
      & \mqty{1&0\\0&1}
    \end{array}
  \,
  \right]
$$
As you can see, this is a diagonal block matrix: a $(4\times 4)$ matrix with $(2\times 2)$ matrices as entries.
The rows and the columns of the $(4\times 4)$ matrix are labelled by^[We always use the [**lexicographic order**](https://en.wikipedia.org/wiki/Lexicographic_order) $00<01<10<11$.] the binary strings $00, 01, 10, 11$, and the $(2\times 2)$ matrices on the diagonal represent operations applied to the qubit in the second register.
Here, all of these matrices on the diagonal are the identity $\id$ except for the $(01,01)$ entry (i.e. the second one), which is the bit-flip $X$.
This is because $f(01)=1$ (and so we want to turn the control value $y=0$ into $y=1$, which is achieved by applying the bit-flip operator), but $f(x)=0$ for all other binary strings $x$ (and so we want to leave the control value $y=0$ as it is).






## More phase kick-back {#more-phase-kick-back}

<div class="video" title="Phase kick-back" data-videoid="iSCYkhkWVjw"></div>

What makes quantum evaluation of Boolean functions really interesting --- and what truly sets it apart from classical evaluation --- is its action on a *superposition of different inputs*.
For example,^[We make two notational simplifications: we usually don't worry about normalisation factors, and we often just write $\sum_x$ to mean $\sum_{x\in\{0,1\}^n}$.]
$$
  \sum_{x}\ket{x}\ket{0}
  \longmapsto
  \sum_{x}\ket{x}\ket{f(x)}
$$
produces $f(x)$ for *all* $x$ in a *single* run.
However, it is more instructive to see the effect of quantum function evaluation when the qubit in the second register is prepared in the state $\ket{-}\coloneqq\frac{1}{\sqrt{2}}(\ket{0}-\ket{1})=H\ket{1}$, since then
$$
  \sum_x\ket{x}\ket{-}
  \longmapsto
  \sum_x (-1)^{f(x)}\ket{x}\ket{-}
$$
(as shown in Figure \@ref(fig:n-equals-3-minus-circuit-diagram)).
In words, whenever $f(x)=1$, the bit flip $X$ is applied to the qubit in the second register.

(ref:n-equals-3-minus-circuit-diagram-caption) Computing some $f\colon\{0,1\}^3\to\{0,1\}$ with the second register in state $\ket{-}$.

```{r n-equals-3-minus-circuit-diagram,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex"),fig.cap='(ref:n-equals-3-minus-circuit-diagram-caption)'}
\begin{equation*}
  \begin{quantikz}
    \lstick[wires=3]{\footnotesize$\sum_x\ket{x}$}
    & \ctrl{3}
    & \qw \rstick[wires=3]{\footnotesize$\sum_x(-1)^{f(x)}\ket{x}$}
  \\[-1em]
    & \ctrl{2}
    & \qw
  \\[-1em]
    & \ctrl{1}
    & \qw
  \\\lstick{$\ket{-}$}
    & \gate{f}
    & \qw \rstick{$\ket{-}$}
  \end{quantikz}
\end{equation*}
```

The reason for defining the state $\ket{-}$ as we do is that it is the eigenstate of $X$ with eigenvalue $-1$, i.e. $X\ket{-}=-\ket{-}$.
So, due the phase kick-back, whenever $f(x)=1$, the phase factor $-1$ appears in front of the corresponding term $\ket{x}$.
As you can see, the second register stays in state $\ket{-}$ all the way through the computation --- it is the first register where things happen.

Let us now see how quantum Boolean function evaluation introduces phase shifts in quantum interference experiments, and how such experiments can be viewed as computations.






## Oracles

The computational power of quantum interference was discovered by counting how many times certain Boolean functions have to be evaluated in order to find the answer to a given problem.
Imagine a "black box" (sometimes also called an **oracle**) that computes some fixed Boolean function, but whose inner workings are unknown to us.
Then imagine that we are in a scenario where we want to learn about some given property of the Boolean function, but we have to "pay" (in energy, time, money, or anything!) for each use (often referred to as a **query**) of the box.
In such a setting, the objective is to minimise number of queries to the oracle while finding out as much information as possible about the function that it computes.
For this purpose, we ignore everything that happens inside the black box: in our rules of the game, the Boolean function evaluation counts as just *one* computational step.






## Deutsch's algorithm {#deutschs-algorithm}

<div class="video" title="Deutsch's algorithm" data-videoid="DyINHZoOcLQ"></div>

We start, once more, with the simplest quantum interference circuit

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \phase{\varphi}
    & \gate{H}
    & \qw \rstick{$\cos\frac{\varphi}{2}\ket{0} - i\sin\frac{\varphi}{2}\ket{1}$}
  \end{quantikz}
\end{equation*}
```

but let's turn this into a black-box scenario, often known as the **binary observable measurement** problem:^[You might recognise this as Exercise \@ref(unknown-phase), and we will return to this problem again in Section \@ref(phase-estimation).]

- we are allowed to prepare the input in any state that we like;
- we are allowed to read the output;
- *but* all we know about the value of $\varphi$ is that it is either $0$ or $\pi$, and we are *not* allowed to "look inside" the phase gate to see which value it is!

With these rules, can we figure out which value $\varphi$ has been set to?

Of course we can --- we're quantum information scientists!

One way of doing it is to prepare the input in the state $\ket{0}$ and check the output: if $\varphi=0$ then the output is always $\ket{0}$, and if $\varphi=\pi$ then it is always $\ket{1}$.
In other words, a *single run* of the interference experiment is sufficient to determine the difference.

The very first quantum algorithm, proposed by [David Deutsch](https://en.wikipedia.org/wiki/David_Deutsch) in 1985, is very much related to this effect, but where the phase setting is determined by the Boolean function evaluation via the phase kick-back.

:::: {.scenario latex=""}
::: {.scenario-title}
*(Global properties of a one-bit function).*
:::
We are presented with an oracle that computes some unknown function $f\colon\{0,1\}\to\{0,1\}$.
Note that there are only four possibilities for what $f$ can be: it could be one of two **constant** functions (i.e. those where $f(0)=f(1)$), or one of two **balanced** functions (i.e. those where $f(0)\neq f(1)$).

|    | $f(0)$ | $f(1)$ |
| -: | :----: | :----: |
| constant | $0$ | $0$ |
| constant | $1$ | $1$ |
| balanced | $0$ | $1$ |
| balanced | $1$ | $0$ |

Our task is to determine, using the fewest queries possible, whether the function computed by the oracle is constant or balanced.
::::

Note that we are *not* asked for the particular values $f(0)$ and $f(1)$, but *only whether the two values are the same or different*.
Classical intuition tells us that we have to evaluate both $f(0)$ and $f(1)$ and compare them, which involves evaluating $f$ *twice*.
But, in the quantum setting, we can solve this problem with a *single* function evaluation, using the following circuit.^[The original version of Deutsch's algorithm provides the correct answer with probability 50%. Here we present a modified/improved version. The more general problem, which deals with unknown functions $f\colon\{0,1\}^n\to\{0,1\}$ for $n\geq1$, is known as the [Deutsch--Jozsa problem](https://en.wikipedia.org/wiki/Deutsch%E2%80%93Jozsa_algorithm).]

:::: {.circuit latex=""}
::: {.circuit-title}
*(Deutsch's algorithm).*
:::
First register: $1$ qubit. Second register: $1$ qubit.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \ctrl{1}
    & \gate{H}
    & \qw \rstick{$\begin{cases}\ket{0}&\mbox{if constant}\\\ket{1}&\mbox{if balanced}\end{cases}$}
  \\\lstick{$\ket{-}$}
    & \qw
    & \gate{f}
    & \qw
    & \qw \rstick{$\ket{-}$}
  \end{quantikz}
\end{equation*}
```
::::

During the function evaluation, the second register "kicks back" the phase factor $(-1)^{f(x)}$ in front of $\ket{x}$, but the state of the second register remains unchanged; the first register is modified as follows:
$$
  \begin{aligned}
    \ket{0}
    &\xmapsto{H}
    \ket{0}+\ket{1}
  \\&\xmapsto{U_f}
    (-1)^{f(0)}\ket{0} + (-1)^{f(1)}\ket{1}
  \\&\quad=
    \ket{0} + (-1)^{f(0)\oplus f(1)}\ket{1}
  \\&\xmapsto{H}
    \ket{f(0)\oplus f(1)}.
  \end{aligned}
$$

The evolution of the first qubit is thus identical to that described by the circuit diagram

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \phase{\varphi}
    & \gate{H}
    & \qw \rstick{$\begin{cases}0&\mbox{if constant}\\1&\mbox{if balanced}\end{cases}$}
  \end{quantikz}
\end{equation*}
```

where the relative phase is $\varphi = (-1)^{f(0)\oplus f(1)}$.
The first qubit ends in state $\ket{0}$ if the function $f$ is constant, and in state $\ket{1}$ if the function is balanced, and the standard measurement distinguishes these two cases with certainty.^[This is also implemented in the [Quantum Flytrap Virtual Lab](https://lab.quantumflytrap.com/lab/deutsch-jozsa?mode=laser).]

But really this is just the binary observable measurement problem in disguise!
Indeed, the fact that quantum Boolean function evaluation of a function $f$ is given by
$$
  \ket{x}\ket{y} \longmapsto \ket{x}\ket{y\oplus f(x)}
$$
means that the unitary $U_f$ has eigenvalues $\pm1$ because it satisfies $U_f^2=\id$, since two consecutive evaluations gives
$$
  \begin{aligned}
  \ket{x}\ket{y}
  \longmapsto& \ket{x}\ket{y\oplus f(x)}
\\\longmapsto& \ket{x}\ket{y\oplus f(x)\oplus f(x)}
\\=& \ket{x}\ket{y}.
  \end{aligned}
$$
So the fact that $1=e^{0}$ and $-1=e^{i\pi}$ means that $U_f$ acts as a phase gate with phase either $0$ or $\pi$.
We will come back to this link between Deutsch's algorithm and binary observable measurement in Section \@ref(phase-estimation).

Deutsch's result laid the foundation for the new field of quantum computation, and was followed by several other quantum algorithms for various problems.
They all seem to rest on the same generic sequence:

- a Hadamard transform;
- function evaluation;
- another Hadamard (or Fourier) transform.^[As explained in Section \@ref(quantum-fourier-transform), the Hadamard transform is a special case of the Fourier transform over the group $\mathbb{Z}_2^n$.]

As we shall see in a moment, in some cases (such as in Grover's search algorithm) this sequence is repeated several times.

Let us now follow a tour through the three early quantum algorithms, where each one offers a higher-order speed-up when compared to their classical analogues than the last: firstly linear, then quadratic, and finally exponential.
After this, we will look at generalising binary observable measurement, the corresponding algorithm analogous to Deutsch's, and how this leads us to arguably the most famous quantum algorithm: Shor's algorithm for prime factorisation.






## The Bernstein--Vazirani algorithm {#the-bernstein-vazirani-algorithm}

<div class="video" title="The Bernstein–Vazirani algorithm" data-videoid="60OHCftlqbA"></div>

:::: {.scenario latex=""}
::: {.scenario-title}
*(Hidden inner-product determination).*
:::
We are presented with an oracle that computes some unknown function $f\colon\{0,1\}^n\to\{0,1\}$, but we are promised that $f$ is of the form
$$
  f(x) = a\cdot x
  \equiv (a_1\cdot x_1) \oplus \ldots \oplus (a_n\cdot x_n)
$$
for some fixed binary string $a=a_1a_2\ldots a_n\in\{0,1\}^n$.

Our task is to determine, using the fewest queries possible, the value of the $n$-bit string $a$.
::::

It's quite easy to see how to do this classically: if we input the value $x=00\ldots010\ldots0$, where the $m$-th bit is a $1$ and all other bits are $0$, then $f(x)$ is simply the $m$-th bit of $a$; after $n$ such calls, we will know every bit value, and thus know $a$.
It is also clear that there *cannot* exist a better classical algorithm: each call to the oracle teaches us exactly one bit of information, and since we must learn $n$ bits, we must query it $n$ times.

In contrast, by running the circuit below, it is possible to determine the value of $a$ with a *single* call to the oracle!

:::: {.circuit latex=""}
::: {.circuit-title}
*(The Bernstein-Vazirani algorithm).*
:::
First register: $n$ qubits. Second register: $1$ qubit.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\newcommand{\vdotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={yshift=0.3em}]{\vdots}}
\begin{equation*}
  \begin{quantikz}
    \lstick[wires=4]{$\ket{0}^{\otimes n}$}
    & \gate{H}
    & \ctrl{4}
    & \gate{H}
    & \qw \rstick{$\ket{a_1}$}
  \\[-1em]
    & \gate{H}
    & \ctrl{3}
    & \gate{H}
    & \qw \rstick{$\ket{a_2}$}
  \\[-1em]
    &&\vdotsgate
  \\[-1em]
    & \gate{H}
    & \ctrl{1}
    & \gate{H}
    & \qw \rstick{$\ket{a_n}$}
  \\[2em]
    \lstick{$\ket{-}$}
    & \qw
    & \gate{f}
    & \qw
    & \qw \rstick{$\ket{-}$}
  \end{quantikz}
\end{equation*}
```
::::

A quick note on notation: the "$\ldots$" in the circuit means "there are more wires here but they are identical (apart from the numbering) to the ones above".
You might also see other notation to denote this, such as

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}^{\otimes n}$}
    &[2mm] \gate{H}\qwbundle{n}
    & \qw
  \end{quantikz}
\end{equation*}
```

or even simply

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}^{\otimes n}$}
    & \gate{H}\qwbundle[alternate]{}
    & \qwbundle[alternate]{}
  \end{quantikz}
\end{equation*}
```

Now, stepping through the execution of the circuit (and ignoring the second register, which, as per usual, remains in the state $\ket{-}$ throughout), we obtain
$$
  \begin{aligned}
    \ket{0^{\otimes n}}
    &\xmapsto{H^{\otimes n}}
    \left(\frac{1}{\sqrt{2}}\right)^n \sum_{x\in\{0,1\}^n} \ket{x}
  \\&\xmapsto{U_f}
    \left(\frac{1}{\sqrt{2}}\right)^n \sum_{x\in\{0,1\}^n} (-1)^{a\cdot x}\ket{x}
  \\&\xmapsto{H^{\otimes n}}
    \left(\frac{1}{\sqrt{2}}\right)^n \sum_{x\in\{0,1\}^n}
    \left[
      (-1)^{a\cdot x} \left(\frac{1}{\sqrt{2}}\right)^n
      \sum_{y\in\{0,1\}^n} (-1)^{y\cdot x} \ket{y}
    \right]
  \\&\quad= \left(\frac{1}{2}\right)^n \sum_{y\in\{0,1\}^n}
    \left[
      \sum_{x\in\{0,1\}^n} (-1)^{(a\oplus y)\cdot x}
    \right]
    \ket{y}
  \end{aligned}
$$
where we write the second Hadamard transform as
$$
  \ket{x}
  \longmapsto
  \left(\frac{1}{\sqrt{2}}\right)^n \sum_{y\in\{0,1\}^n} (-1)^{y\cdot x}\ket{y}.
$$

To see that this output is indeed equal to $\ket{a}$, we can use the fact^[**Exercise.** Prove this!] that, for any $y\in\{0,1\}^n$,
$$
  \sum_{x\in\{0,1\}^n} (-1)^{x\cdot y}
  = \begin{cases}
    0 &\text{if $y\neq0$;}
  \\2^n &\text{if $y=0$}
  \end{cases}
$$
which, in our case, tells us that
$$
  \sum_{x\in\{0,1\}^n} (-1)^{(a\oplus y)\cdot x}
  = \begin{cases}
    0 &\text{if $y\neq a$;}
  \\2^n &\text{if $y=a$.}
  \end{cases}
$$

In other words, if you take the sum over $x$, then all the terms always cancel out *unless* $a\oplus y = 00\ldots0$, but this happens *if and only if* $y=a$.
Then the standard bit-by-bit measurement of the first register gives the value of $a$ and solves the problem with a single call to the oracle.

Alternatively, if you don't immediately see how this sum works for $z\neq a$ (where we write $\ket{z}$ to mean the output), you can first calculate the probability that the output is $z=a$.
In this case it is easy to see that the sum is $2^n$, and that in the final state $\sum_z\lambda_z\ket{z}$ the term $z=a$ has amplitude $1$.
Thus, by normalisation, all the other terms must be equal to $0$.


## Grover's search algorithm

The next algorithm we will study aims to solve the problem of searching for a specific item in an *unsorted* database.
Think about an old-fashioned phone book: the entries are typically sorted alphabetically, by the name of the person that you want to find.
However, what if you were in the opposite situation: you had a phone number and wanted to find the corresponding person's name?
The phone book is not sorted in that way, and to find the number (and hence name) with, say, 50% probability, you would need to search through, on average, 50% of the entries.
Needless to say, in a large phone book this would take a long time.

While this might seem like a rather contrived problem (a computer database *should* always maintain an index on any searchable field), many problems in computer science can be cast in this form, i.e. that of an **unstructured search**.

:::: {.scenario latex=""}
::: {.scenario-title}
*(Unstructured search).*
:::
We are presented with an oracle that computes some unknown function $f\colon\{0,1\}^n\to\{0,1\}$.

Our task is to find, using the fewest queries possible, an input $x\in\{0,1\}^n$ such that $f(x)=1$.
::::

Suppose that we know that, amongst the $N=2^n$ binary strings, there are $M\ll N$ which are **tagged**, i.e. strings on which $f$ evaluates to $1$.
Since there is no structure in the database, any classical search requires around $N/M$ steps, i.e. the function $f$ must be evaluated roughly $N/M$ times.

In contrast, there is a quantum search algorithm, implemented by the circuit below, that was proposed in 1996 by [Lov Grover](https://en.wikipedia.org/wiki/Lov_Grover), and which requires only roughly $\sqrt{N/M}$ steps.

:::: {.circuit latex=""}
::: {.circuit-title}
*(Grover's search).*
:::
First register: $n$ qubits. Second register: $1$ qubit.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\newcommand{\vdotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={yshift=0.3em}]{\vdots}}
\begin{equation*}
  \begin{quantikz}
    \lstick[wires=4]{$\ket{0}^{\otimes n}$}
    & \gate{H}
    & \qw
    & \ctrl{4}
      \gategroup[wires=5,steps=4,background,style={rounded corners,dashed,ultra thick,inner sep=0.5em},label style={yshift=0.4em,align=center}]{Grover iteration operator $G$\\(repeat $O(2^{n/2})$ times)}
    & \gate{H}
    & \ctrl{4}
    & \gate{H}
    & \qw
    & \qw \rstick{$\ket{z_1}$}
  \\[-1em]
    & \gate{H}
    & \qw
    & \ctrl{3}
    & \gate{H}
    & \ctrl{3}
    & \gate{H}
    & \qw
    & \qw \rstick{$\ket{z_2}$}
  \\[-1em]
    &&&\vdotsgate
    &&\vdotsgate
  \\[-1em]
    & \gate{H}
    & \qw
    & \ctrl{1}
    & \gate{H}
    & \ctrl{1}
    & \gate{H}
    & \qw
    & \qw \rstick{$\ket{z_n}$}
  \\[2em]
    \lstick{$\ket{-}$}
    & \qw
    & \qw
    & \gate{f}
    & \qw
    & \gate{f_0}
    & \qw
    & \qw
    & \qw \rstick{$\ket{-}$}
  \end{quantikz}
\end{equation*}
```

where $f_0$ tags the binary string consisting of $n$ zeros:
$$
  f_0(x)
  = \begin{cases}
    1 &\text{if $x=00\ldots0$;}
  \\0 &\text{otherwise.}
  \end{cases}
$$
::::

Yet again, we can recognise the typical Hadamard, function evaluation, Hadamard sequence, and yet again we can see that the second register (the bottom qubit, in state $\ket{-}$) plays an auxiliary role: the real action takes place in the first register.
However, unlike the previous algorithms, a single call to the oracle does not do very much, and we have to build up the quantum interference in the first register through repeated calls to the oracle (without any intermediate measurements!).

Here, the basic step is the **Grover iteration operator $G$**, which is the boxed part of the circuit that we repeat over and over:
$$
  G = (H^{\otimes n}\otimes\id) U_{f_0}(H^{\otimes n}\otimes\id) U_f.
$$

After^[Recall the big-$O$ notation introduced in Exercise \@ref(big-o).] $O(2^{n/2})$ applications of $G$, we measure the first register bit-by-bit and obtain the value of $\ket{z}=\ket{z_1z_2\ldots z_n}$, which is such that, *with "high" probability*, $f(z)=1$.
In order to actually *see* how this algorithm works, and to justify our use of the phrase "with high probability", we can take a more geometric approach.

First, we define two orthonormal vectors in the Hilbert space describing the first register:^[Once again, we shall completely ignore the second register from now on, since all the interesting stuff happens in the first.]
$$
  \begin{aligned}
    \ket{a}
    &= \frac{1}{\sqrt{N-M}} \sum_{x\in f^{-1}(0)} \!\!\!\!\ket{x}
  \\\ket{b}
    &= \frac{1}{\sqrt{M}} \sum_{x\in f^{-1}(1)} \!\!\!\!\ket{x}
  \end{aligned}
$$
where $f^{-1}(i)\coloneqq\{x\in\{0,1\}^n\mid f(x)=i\}$ is the **preimage** of $i$.
Since these vectors are orthonormal, they are, in particular, linearly independent, and so their span is a two-dimensional subspace --- this is the subspace in which our search will take place.

Using the fact that $f^{-1}(0)$ and $f^{-1}(1)$ form a **partition**^[A **partition** of a set $X$ is a collection of disjoint subsets $X_1,\ldots,X_m\subseteq X$ whose union is all of $X$, i.e. $X_i\cap X_j=\varnothing$ for all $i\neq j$, and $X_1\cup\ldots\cup X_m=X$.] of the space $\{0,1\}^n$, we see that the two-dimensional span of $\ket{a}$ and $\ket{b}$ contains, in particular, the equally-weighted superposition $\ket{s}=H^{\otimes n}\ket{0^{\otimes n}}$ of all binary strings of length $n$:
$$
  \begin{aligned}
    \ket{s}
    &= \frac{1}{\sqrt{N}}\sum_{x\in\{0,1\}^n}\ket{x}
  \\&= \frac{1}{\sqrt{N}}\sum_{x\in f^{-1}(0)}\ket{x} + \frac{1}{\sqrt{N}}\sum_{x\in f^{-1}(1)}\ket{x}
  \\&= \sqrt{\frac{N-M}{N}}\ket{a} + \sqrt{\frac{M}{N}}\ket{b}
  \\&= (\cos\theta)\ket{a} + (\sin\theta)\ket{b}
  \end{aligned}
$$
where we use the fact that
$$
  \sqrt{\frac{N-M}{N}}^2 + \sqrt{\frac{M}{N}}^2
  = 1
  = \sin^2\theta+\cos^2\theta
$$
to parametrise $\sqrt{\frac{N-M}{N}}$ as $\cos\theta$ and $\sqrt{\frac{M}{N}}$ as $\sin\theta$ (with $\theta\approx\sqrt{\frac{M}{N}}$, since $N\gg M$).

The state $\ket{s}$ is our starting input for our sequence of Grover iterations, and we will show that applying $G$, when restricting to the plane spanned by $\ket{a}$ and $\ket{b}$, amounts to applying a rotation by angle $2\theta$.
Grover's search algorithm can then be understood as a sequence of rotations which take the input state $\ket{s}$ towards the target state $\ket{b}$.

To see this, note that the unitary transformation induced by the oracle
$$
  f\colon \ket{x} \mapsto (-1)^{f(x)}\ket{x}
$$
can be written as^[To prove this, it suffices to check that these two transformations agree on the standard basis; since $f^{-1}(0)$ and $f^{-1}(1)$ form a partition, we know that any element (and, in particular, any basis element) is either in the preimage of $0$ or of $1$; if it is in the former, then $I_a$ acts as the identity; if the latter, then $I_a$ acts as $-\id$.]
$$
  I_a \coloneqq 2\proj{a}-\id
$$
which we can interpret as a reflection through the $\ket{a}$-axis: we see that
$$
  \begin{aligned}
    I_a\ket{a}
    &= 2\proj{a}\ket{a}-\ket{a}
  \\&= 2\ket{a}-\ket{a}
  \\&= \ket{a}
  \\I_a\ket{b}
    &= 2\proj{a}\ket{b}-\ket{b}
  \\&= -\ket{b}
  \end{aligned}
$$
and since $-\ket{b}$ is a vector that points in the opposite direction from $\ket{b}$, it must be a reflection; since $-\ket{b}$ is still orthogonal to $\ket{a}$, the reflection must be in the $\ket{a}$-axis.

Some further algebraic manipulation shows that $I_a = 2\proj{a}-\id = \id-2\proj{b}$.
Now, in particular, evaluation of $f_0$ can be written as $2\proj{0}-\id$, and thus thought of as a reflection through the $\ket{0}$-axis.
If we sandwich $f_0$ in between two Hadamards then we obtain $I_s = 2\proj{s}-\id$, which is reflection through the $\ket{s}$-axis.
By definition then, the Grover iteration operator $G$ is the composition
$$
  G = I_s I_a.
$$

Now recall the purely geometric fact that if we have two intersecting lines $L_1$ and $L_2$ in two-dimensional Euclidean space, meeting with angle $\alpha$, then reflecting an object through $L_1$ and then reflecting the resulting image through $L_2$ is the same as simply rotating the original object around the point of intersection $L_1\cap L_2$ by an angle of $2\alpha$.

The angle between $\ket{a}$ and $\ket{s}$ is $\theta$, so each time $G$ is applied the vector is rotated (around the origin) by an angle of $2\theta$ towards the $\ket{b}$-axis.
All that remains to do is to just choose the "right" number $r$ of steps such that we end up as close to the $\ket{b}$-axis as possible.
The state $\ket{s}$ starts at angle $\theta$ to $\ket{a}$, and we should perform our final (and only) measurement when this angle is $\pi/2$, i.e. when $(2r+1)\theta = \pi/2$, which gives
$$
  r \approx \frac{\pi}{4}\sqrt{\frac{N}{M}}.
$$

```{r grover-search-geometrically,engine='tikz',fig.cap='Understanding the Grover search algorithm geometrically.',fig.width=6}
\definecolor{primary}{RGB}{177,98,78}
\definecolor{secondary}{RGB}{91,132,177}
\usetikzlibrary{arrows.meta}
\newcommand{\ket}[1]{|#1\rangle}
\begin{tikzpicture}[scale=1.4]
  \begin{scope}
    \draw (0,0) circle (1cm);
    \draw [->] (0,-1.4) to (0,1.4) node [above] {$\ket{b}$};
    \draw [->] (-1.4,0) to (1.4,0) node [right] {$\ket{a}$};
    %
    \draw [thin] (0,0) to (0:0.7) to [bend right,dashed] (20:0.7) to cycle;
    \node at (10:0.6) {\footnotesize$\theta$};
    \draw [-Latex,primary] (0,0) to (20:1) node [right] {$\ket{s}$};
  \end{scope}
  \node at (0,-1.7) {the initial state};
  \begin{scope}[shift={(4,0)}]
    \draw (0,0) circle (1cm);
    \draw [->] (0,-1.4) to (0,1.4) node [above] {$\ket{b}$};
    \draw [->] (-1.4,0) to (1.4,0) node [right] {$\ket{a}$};
    %
    \draw [thin] (0,0) to (-20:0.7) to [bend right=1.8cm] (60:0.7) to cycle;
    \node at (-10:0.6) {\footnotesize$\theta$};
    \node at (10:0.6) {\footnotesize$\theta$};
    \node at (40:0.55) {\footnotesize$2\theta$};
    \draw [-Latex,primary] (0,0) to (20:1) node [right] {$\ket{s}$};
    \draw [-Latex,secondary] (0,0) to (60:1) node [above right] {$I_sI_a\ket{s}$};
    \draw [-Latex,secondary] (0,0) to (-20:1) node [right] {$I_a\ket{s}$};
  \end{scope}
  \node at (4,-1.7) {applying $G=I_sI_a$};
\end{tikzpicture}
```

So the quantum algorithm searches an unsorted list of $N$ items in roughly^[Recall that $M\ll N$, so $\sqrt{\frac{N}{M}}\approx\sqrt{N}$.] $\sqrt{N}$ steps: it offers a *quadratic* speed-up when compared to classical search, which can be of immense practical importance.
For example, cracking some of the more popular modern ciphers, such as [AES](https://en.wikipedia.org/wiki/Advanced_Encryption_Standard), essentially requires a search among *many* binary strings (called **keys**).
If these can be checked at a rate of, say, one million keys per second, then a classical computer would need over a thousand years to find the correct key, while a quantum computer using Grover's algorithm would find it in less than four minutes.



## Simon's algorithm

<div class="video" title="Simon's algorithm" data-videoid="sxIdRKTdRRU"></div>

Here we will see the simplest quantum algorithm that offers an *exponential* speed-up when compared to the best possible classical algorithm.

:::: {.scenario latex=""}
::: {.scenario-title}
*(Hidden binary-addition determination).*
:::
We are presented with an oracle that computes some unknown function $f\colon\{0,1\}^n\to\{0,1\}^n$, but we are promised that $f$ satisfies, for all $x\in\{0,1\}^n$,
$$
  f(x) = f(x\oplus s)
$$
for some fixed $s\in\{0,1\}^n$, which we call the **period** of $f$.
(We assume that $s$ is *not* the string of $n$ zeros, otherwise the problem becomes trivial.)

Our task is to determine, using the fewest queries possible, the value of the $n$-bit string $s$.
::::

Note that asking for $f$ to be periodic is equivalent to asking that $f$ be **two-to-one**: for any $y\in\{0,1\}^n$ such that there exists some $x\in\{0,1\}^n$ with $f(x)=y$, there exists *exactly one other* $x'\neq x$ such that $f(x')=y$ as well.

Classically, this problem is exponentially hard.
We will not go through a detailed proof of this fact, but the intuition is reasonably simple: since there is no structure in the function $f$ that would help us find its period $s$, the best we can do is evaluate $f$ on random inputs and hope that we find some distinct $x$ and $x'$ such that $f(x)=f(x')$, and then we know that $s=x\oplus y$.
After having made $m$ queries to the oracle, we have a list of $m$-many tuples $(x,f(x))$; there are $m(m-1)/2$ possible pairs which could match within this list, and the probability that a randomly chosen pair match is $1/2^{n-1}$.
This means that the probability of there being at least one matching pair within the list of $m$ tuples is less than $m^2/2^n$.
This means that the chance of finding a matching pair is negligible if the oracle is queried on fewer than $m=\sqrt{2^n}$ inputs.

The quantum case, on the other hand, gives a result (again, with "high" probability) within a *linear* number of steps.
The circuit that solves this problem, shown below, has a familiar Hadamard--function--Hadamard structure, but the second register has now been expanded to $n$ qubits.

:::: {.circuit latex=""}
::: {.circuit-title}
*(Simon's problem).*
:::
First register: $n$ qubits. Second register: $n$ qubits.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\newcommand{\vdotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={yshift=0.3em}]{\vdots}}
\begin{equation*}
    \begin{quantikz}
      \lstick[wires=4]{$\ket{0}^{\otimes n}$}
      & \gate{H}
      & \ctrl{4}
      & \gate{H}
      & \qw
    \\[-1em]
      & \gate{H}
      & \ctrl{3}
      & \gate{H}
      & \qw
    \\[-1em]
      &&\vdotsgate
    \\[-1em]
      & \gate{H}
      & \ctrl{1}
      & \gate{H}
      & \qw
    \\[2em]
      \lstick[wires=4]{$\ket{0}^{\otimes n}$}
      & \qw
      & \gate[wires=4,nwires={3}]{f}
      & \qw
      & \qw
    \\[-1em]
      & \qw
      && \qw
      & \qw
    \\[-1em]
      &\vdotsgate
      &&\vdotsgate
    \\[-2em]
      & \qw
      && \qw
      & \qw
    \end{quantikz}
\end{equation*}
```
::::

This time, let's follow the evolution of *both* registers throughout this circuit.
We start off by preparing the equally-weighted superposition of all $n$-bit strings with the first Hadamard, and then query the oracle:
$$
  \begin{aligned}
    \ket{0^{\otimes n}}\ket{0^{\otimes n}}
    &\xmapsto{H^{\otimes n}\otimes\id}
    \frac{1}{\sqrt{2^n}} \sum_x \ket{x}\ket{0^{\otimes n}}
  \\&\xmapsto{U_f}
    \frac{1}{\sqrt{2^n}} \sum_x \ket{x}\ket{f(x)}.
  \end{aligned}
$$
The second Hadamard transform on the first register then yields the final output state:
$$
  \frac{1}{2^n} \sum_{x,y} (-1)^{x\cdot y} \ket{y}\ket{f(x)}.
\tag{$\ddagger$}
$$

But if we measure^[As we shall see in a moment, the actual measurement on the second register is not actually necessary.] the second register *before* applying the second Hadamard transform to the first register, we obtain one of the $2^{n-1}$ possible values of $f(x)$, each equally likely.
Let's study the implications of this.

Suppose that the outcome of the measurement is $f(a)$ for some $a\in\{0,1\}^n$.
Given that both $a$ and $a\oplus s$ are mapped to $f(a)$ by $f$, the first register then collapses to the state
$$
  \frac{1}{\sqrt{2}}\big( \ket{a} + \ket{a\oplus s} \big).
$$
The subsequent Hadamard transform on the first register then gives us the final state^[We write $s^\perp$ to mean the set of all $y\in\{0,1\}^n$ such that $y\cdot s=0$.]
$$
  \begin{aligned}
    &\frac{1}{\sqrt{2^{n+1}}} \sum_y \Big((-1)^{a\cdot y}+(-1)^{(a\oplus s)\cdot y}\Big) \ket{y}\ket{f(a)}
  \\= &\frac{1}{\sqrt{2^{n+1}}} \sum_y (-1)^{a\cdot y}
      \Big( 1 + (-1)^{s\cdot y} \Big) \ket{y}\ket{f(a)}
  \\= &\frac{1}{\sqrt{2^{n-1}}} \sum_{y\in s^\perp}
      (-1)^{a\cdot y} \ket{y}\ket{f(a)}
  \end{aligned}
$$
where we have used the fact that $(a\oplus s)\cdot y = (a\cdot y)\oplus(s\cdot y)$, and that $1+(-1)^{s\cdot y}$ can have only two values: either $2$ (when $s\cdot y = 0$) or $0$ (when $s\cdot y = 1$).
Now we finally measure the first register: the outcome is selected at random from all possible values of $y$ such that $a\cdot y = 0$, each occurring with probability $1/(2^{n-1})$.

In fact, we did not have to measure the second register at all: it was a mathematical shortcut, simply taken for pedagogical purposes.
Instead of collapsing the state to just one term in a superposition, we can express Equation ($\ddagger$) as
$$
  \begin{aligned}
    &\frac{1}{2^n} \sum_{y,f(a)}
    \Big( (-1)^{a\cdot y} + (-1)^{(a\oplus s)\cdot y} \Big) \ket{y}\ket{f(a)}
  \\= &\frac{1}{2^n} \sum_{y,f(a)} (-1)^{a\cdot y}
    \Big( 1 + (-1)^{s\cdot y} \Big) \ket{y}\ket{f(a)}
  \end{aligned}
$$
where the summation over $f(a)$ means summing over all binary strings in the image of $f$, which is a convenient shorthand for the more complicated^[Another (complicated sounding) way of expressing this sum is by choice of a **section** of $f$, i.e. picking *one* element $a_z$ in *each* preimage $f^{-1}(z)$ is equivalent to defining a function $a\colon\{0,1\}^n\to\{0,1\}^n$ by $a\mapsto a_z$ which satisfies $(a\circ f)(z)=z$, or $a\circ f=\id$.] formal statement: for all $z\in\{0,1\}^n$, since our function $f$ is two-to-one, we know that the preimage $f^{-1}(z)$ has either *two* elements or *none*; if it's the latter, then we don't include $z$ in our sum; if it's the former, we *pick one* and call it $a$ (knowing that the other, by the setup, must be $a\oplus s$, which you can see appearing in the first equation above).
With this, the final output of the algorithm is
$$
  \frac{1}{2^{n-1}} \sum_{y\in s^\perp} \ket{y}
    \sum_{f(a)} (-1)^{a\cdot y} \ket{f(a)}
$$
and, again, the measurement outcome is selected at random from all possible values of $y$ such that $s\cdot y=0$.

We are not quite done yet: we cannot infer $s$ from a _single_ output $y$.
However, once we have found $n-1$ linearly independent strings^[It is important to note that we are talking about the pure "abstract" *binary strings*, and not the corresponding states in the Hilbert space. Concretely, this means that **linearly independent** here means "no string in the set $\{y_1,\ldots,y_n\}$ can be expressed as the bitwise sum of some other strings in this set". That is, we are working with the $\mathbb{Z}/2\mathbb{Z}$-vector space of strings, not the $\mathbb{C}$-vector space of states!] $y_1,y_2,\ldots,y_{n-1}$, we can solve the $n-1$ equations
$$
  \left\{
  \begin{aligned}
    s\cdot y_1 &= 0
  \\s\cdot y_2 &= 0
  \\&\,\,\,\vdots
  \\s\cdot y_{n-1} &= 0
  \end{aligned}
  \right\}
$$
to determine a unique value of $s$.
(Note that we only need $n-1$ values, and not $n$, because $s=0$ will always be a solution, but we have explicitly assumed that this is not the case in our statement of the scenario, and so it suffices to narrow down the space of possible solutions to consist of *two* elements, since then we know that we can just take the non-zero one.)

So we run this algorithm repeatedly, each time obtaining another value of $y$ that satisfies $s\cdot y = 0$.
Every time we find some new value of $y$ that is linearly independent of all previous ones, we can discard half the potential candidates for $s$.

(ref:simons-diagramatically-caption) Picture all possible binary strings as dots, but with the string $s$ denoted by a star. Every linearly independent $y_{k+1}$ lets us "zoom in" twice as close towards $s$.

```{r simons-diagramatically,engine='tikz',fig.width=6,fig.cap='(ref:simons-diagramatically-caption)'}
\definecolor{primary}{RGB}{177,98,78}
\definecolor{secondary}{RGB}{91,132,177}
\tikzset{dot/.style={fill,shape=circle,minimum size=5pt,inner sep=0pt}}
\usetikzlibrary{shapes}
\begin{tikzpicture}[scale=1.1]
  \begin{scope}
    \draw (0,0) circle (1.1cm);
    \node [draw,primary,star,star points=5,fill=white,inner sep=0.02cm] at (-22.5:0.7) {$s$};
    \foreach \a in {22.5,67.5,...,292.5}
      \node [dot,fill=primary] at (\a:0.7) {};
  \end{scope}
  \begin{scope}[shift={(3,0)}]
    \draw (0,0) circle (1.1cm);
    \draw [dashed,ultra thick,secondary] (0,-1.3) to (0,1.3) node [above] {$y_1$};
    \node [draw,primary,star,star points=5,fill=white,inner sep=0.02cm] at (-22.5:0.7) {$s$};
    \foreach \a in {22.5,67.5,292.5}
      \node [dot,fill=primary] at (\a:0.7) {};
    \foreach \a in {112.5,157.5,...,247.5}
      \node [dot,fill=black!40] at (\a:0.7){};
  \end{scope}
  \begin{scope}[shift={(6,0)}]
    \draw (0,0) circle (1.1cm);
    \draw [dashed,ultra thick,secondary] (0,-1.3) to (0,1.3) node [above] {$y_1$};
    \draw [dashed,ultra thick,secondary] (-1.3,0) to (1.3,0) node [right] {$y_2$};
    \node [draw,primary,star,star points=5,fill=white,inner sep=0.02cm] at (-22.5:0.7) {$s$};
    \foreach \a in {22.5,67.5,...,247.5}
      \node [dot,fill=black!40] at (\a:0.7) {};
    \node [dot,fill=primary] at (292.5:0.7) {};
  \end{scope}
\end{tikzpicture}
```

Now, the probability that $(n-1)$-many outputs $y_1,\ldots,y_{n-1}$ are linearly independent is
$$
  \left( 1-\frac{1}{2^{n-1}} \right)
  \left( 1-\frac{1}{2^{n-2}} \right)
  \ldots
  \left( 1-\frac{1}{2} \right).
\tag{$\circledast$}
$$
To see this, suppose that we have $k$ linearly independent binary strings $y_1,\ldots,y_k$.
Then these strings span a subspace with $2^k$ elements, consisting of all binary strings of the form $\bigoplus_{i=1}^k b_i y_i$, where $b_1,\ldots,b_k\in\{0,1\}$.
Now suppose we obtain some $y_{k+1}$.
It will be linearly independent from the $y_1,\ldots,y_k$ if and only if it lies *outside* the subspace spanned by the $y_1,\ldots,y_k$, which occurs with probability $1-(2^k)/(2^n)$.

We can bound Equation ($\circledast$) from below:^[Use the inequality $$\begin{aligned}(1-x)(1-y)&= 1 - x - y - xy\\&\geq 1 - (x+y)\end{aligned}$$ which holds for any $0<x,y<1$.]
the probability of obtaining a linearly independent set $\{y_1,\ldots,y_{n-1}\}$ by running the algorithm $n-1$ times (i.e. not having to discard any values and run again) is
$$
  \prod_{k=1}^{n-1}
  \left(
    1-\frac{1}{2^k}
  \right)
  \geq
  \left[
    1 -
    \left(
      \frac{1}{2^{n-1}} + \frac{1}{2^{n-2}} + \ldots + \frac{1}{4}
    \right)
  \right]
  \cdot \frac{1}{2}
  >
  \frac{1}{4}.
$$

We conclude that we can determine $s$ with some constant probability of error after repeating the algorithm $O(n)$ times.
The exponential separation that this algorithm demonstrates between quantum and classical highlights the vast potential of a quantum computer to speed up function evaluation.





## Phase estimation {#phase-estimation}

In Section \@ref(deutschs-algorithm), we took the problem of binary observable measurement and at Deutsch's algorithm, which uses quantum Boolean function evaluation as the controlled-$U$ gate.
Now we're going to generalise the binary observable measurement problem to other controlled-$U$ gates, trying to deduce the value of an unknown phase that can be a lot more general than simply $0$ or $\pi$. Afterwards, we'll explain how this is the first step towards Shor's algorithm for prime factorisation.

Let's start by rephrasing the binary observable measurement problem in terms of **phase estimation** (or **eigenvalue estimation**).
Recall the problem: we are handed some phase gate of unknown phase $\varphi$, but we are promised that $\varphi$ is either $0$ or $\pi$; we want to figure out which one it is by running a simple circuit one time.
But this is entirely equivalent to having access to an oracle that computes a controlled-$U$ gate, along with having an eigenstate $\ket{u}$ with eigenvalue $\pm1$ (i.e. $e^{i\varphi}$ for $\varphi\in\{0,\pi\}$), since this acts as a phase gate of phase $0$ or $\pi$ (depending on whether the eigenvalue is $+1$ or $-1$) on the first register when we plug in $\ket{u}$ to the second register.
This is just phase kick-back again!
So let's state the problem this way, and use the solution that we have already described in Section \@ref(deutschs-algorithm) (which is really the circuit that we saw all the way back in Section \@ref(phase-kick-back)).

:::: {.scenario latex=""}
::: {.scenario-title}
*(Binary phase estimation).*
:::
We are presented with an oracle that computes a controlled-$U$ gate, along with an eigenstate $\ket{u}$ of $U$ with eigenvalue $e^{i\varphi}$
We are promised that $\varphi$ is either $0$ or $\pi$.

Our task is to determine, using the fewest queries possible, the value of $\varphi$.
::::

:::: {.circuit latex=""}
::: {.circuit-title}
*(Controlled-$U$ interference).*
:::
First register: $1$ qubit. Second register: $1$ qubit.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \ctrl{1}
    & \gate{H}
    & \qw \rstick{$\begin{cases}\ket{0}&\mbox{if }\varphi=0\\\ket{1}&\mbox{if }\varphi=\pi\end{cases}$}
  \\\lstick{$\ket{u}$}
    & \qw
    & \gate{U}
    & \qw
    & \qw \rstick{$\ket{u}$}
  \end{quantikz}
\end{equation*}
```
::::

So we see that the eigenvalues being $\pm1$, or, equivalently, $\varphi$ being $0$ or $\pi$, is fundamental to this story --- this is what we want to generalise, so that we can deal with more values^[We won't be able to find the value of completely arbitrary phases exactly, only those of a certain rational form. However, we will talk about the problem of arbitrary phases in Section \@ref(approximate-phase-estimation).] of $\varphi$.

:::: {.scenario latex=""}
::: {.scenario-title}
*(Phase estimation).*
:::
We are presented with an oracle that computes a controlled-$U$ gate, along with an eigenstate $\ket{u}$ of $U$ with eigenvalue $e^{i\varphi}$.
We are promised that $\varphi$ is of the form
$$
  \varphi
  = 2\pi\frac{m}{2^n}
$$
for some integers $m$ and $n$.

Our task is to determine, using the fewest queries possible, the value of ${\varphi}/{2\pi}={m}/{2^n}$
::::

Let's work our way up to finding a circuit to solve this problem, starting with the following simple mathematical observation.
Since $e^{i\varphi}=e^{i(\varphi+2\pi)}$, we can assume $m$ to be an integer between $0$ and $2^n-1$, which means that it has a binary representation of the form
$$
  m
  = \sum_{i=1}^n m_i 2^{n-i}
$$
for $m_i\in\{0,1\}$, and so
$$
  \varphi
  = 2\pi\sum_{i=1}^n m_i 2^{-i}
$$

This helps us rephrase the problem as "find the values of the $m_i$ for $i=1,\ldots,t$".

Now, rather than tackling the full problem, let's try a much smaller modification of the original binary observable measurement problem: we have the same setup --- a phase gate of unknown phase $\varphi$ --- but this time we are promised that $\varphi$ is either $0$ or $\pi/2$ (instead of $0$ or $\pi$).
How can we adapt our original solution to find the phase value?

After some thought, you might see the trick: if we can apply the phase gate *twice in a row*, then we reduce the problem to the one we have already solved.
Indeed, if $\varphi=0$ then repeating it twice is the same as simply applying it once; if $\varphi=\pi/2$ then repeating it twice is the same as simply applying a phase gate of phase $\pi$.

Iterating this idea leads us to the first step of solving the general phase estimation problem, since if $U\ket{u}=e^{i\varphi}\ket{u}$ then $U^{2^{n-1}}\ket{u}=e^{2^{n-1}i\varphi}\ket{u}$.
But
$$
  \begin{aligned}
    2^{n-1}\varphi
    &= 2^n\pi\frac{m}{2^n}
  \\&= \pi m
  \\&= \pi\sum_{i=1}^n m_i 2^{n-i}
  \\&= \pi m_n + \pi\underbrace{\sum_{i=1}^{n-1}m_i2^{n-i}}_{\text{all divisible by }2}
  \end{aligned}
$$
and so, by the $2\pi$-periodicity of $e^{ix}$, we see that
$$
  U^{2^{n-1}}\ket{u}
  = e^{m_n\pi i}\ket{u}
$$
and now we're happy: $m_n\in\{0,1\}$, so this is an eigenvalue of $U^{2^{n-1}}$ with phase $0$ or $\pi$, reducing us to the original binary observable measurement problem that we have already solved by replacing $U$ with $U^{2^{n-1}}$.
We have found the $n$-th bit $m_n$!

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \ctrl{1}
    & \gate{H}
    & \qw \rstick{$\ket{m_n}$}
  \\\lstick{$\ket{u}$}
    & \qw
    & \gate{U^{2^{n-1}}}
    & \qw
    & \qw \rstick{$\ket{u}$}
  \end{quantikz}
\end{equation*}
```

We've dealt with shrinking^[Any interval $[a,b]$ in $\mathbb{R}$ can be turned into any other interval $[c,d]$ by a single shrink (or scale) and a single shift: multiply $a$ and $b$ by $(d-c)/(b-a)$ to get $a'$ and $b'$, and then add $c-a'$ to both $a'$ and $b'$.] the distance between the two possible phases (if we have $0$ and $\pi/k$ for some $k\in\mathbb{N}$ then we simply replace $U$ by $U^k$), so now let's look at the other way of modifying a pair of phases: shifting.
We have the same setup --- a phase gate of unknown phase $\varphi$ --- but this time we are promised that $\varphi$ is either $-\pi/2$ or $\pi/2$ (instead of $0$ or $\pi$).
How can we adapt our original solution to find the phase value?

This time we just need to add $\pi/2$ to each of the phases, since then we'd end up back at $0$ and $\pi$.
This means that we simply need to precede (or follow) the phase gate by the $\pi/4$-phase gate^[Refer to Section \@ref(phase-gates-galore) for an explanation of this confusing terminology.] $S=\left[\begin{smallmatrix}1&0\\0&i\end{smallmatrix}\right]$.

We already know how to find the value of $m_n$, so let's use this to now find the value of $m_{n-1}$.
If we just try the same trick as before then we run into problems: it's true that $U^{2^{n-2}}\ket{u}=e^{2^{n-2}\varphi}\ket{u}$, but
$$
  2^{n-2}\varphi
  \equiv \pi m_{n-1}+\pi m_n/2 \mod2\pi
$$
so we have an annoying phase of $\pi m_n/2$ in the way.
However, we know that all we have to do now is to compensate for this by adding another phase gate:

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \ctrl{1}
    & \phase{{-\pi m_n}/{2}}
    & \gate{H}
    & \qw \rstick{$\ket{m_{n-1}}$}
  \\\lstick{$\ket{u}$}
    & \qw
    & \gate{U^{2^{n-2}}}
    & \qw
    & \qw
    & \qw \rstick{$\ket{u}$}
  \end{quantikz}
\end{equation*}
```

Once we've got the value of $m_{n-1}$ from this circuit, then we can simply apply the same idea to find $m_{n-2}$, $m_{n-3}$, and so on, until we have all the $m_i$ for $i=1,\ldots,n$.
For example, the circuit to determine $m_{n-3}$ (given than we already know $m_n$, $m_{n-1}$, and $m_{n-2}$) is

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\begin{equation*}
  \begin{quantikz}
    \lstick{$\ket{0}$}
    & \gate{H}
    & \ctrl{1}
    & \phase{{-\pi m_n}/{8}}
    & \phase{{-\pi m_{n-1}}/{4}}
    & \phase{{-\pi m_{n-2}}/{2}}
    & \gate{H}
    & \qw \rstick{$\ket{m_{n-3}}$}
  \\\lstick{$\ket{u}$}
    & \qw
    & \gate{U^{2^{n-3}}}
    & \qw
    & \qw
    & \qw
    & \qw \rstick{$\ket{u}$}
  \end{quantikz}
\end{equation*}
```

The entire process for determining the whole string $m_1\ldots m_n$, solving the scenario, can be written in a form that looks exactly how we expect: Hadamard--phase--Hadamard, modulo some subtle changes, which we will explain.

:::: {.circuit latex=""}
::: {.circuit-title}
*(Controlled-$U$ interference).*
:::
First register: $n$ qubits. Second register: $1$ qubit.

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\newcommand{\vdotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={yshift=0.3em}]{\vdots}}
\begin{equation*}
  \begin{quantikz}
    \lstick[wires=4]{$\ket{0}^{\otimes n}$}
    & \gate{H}
    & \ctrl{4}
    & \gate[wires=4,nwires={3}]{FT^\dagger}
    & \qw \rstick{$\ket{m_1}$}
  \\[-1em]
    & \gate{H}
    & \ctrl{3}
    & \qw
    & \qw \rstick{$\ket{m_2}$}
  \\[-1em]
    &&\vdotsgate
  \\[-1em]
    & \gate{H}
    & \ctrl{1}
    & \qw
    & \qw \rstick{$\ket{m_n}$}
  \\[2em]
    \lstick{$\ket{u}$}
    & \qw
    & \gate{U^\triangle}
    & \qw
    & \qw \rstick{$\ket{u}$}
  \end{quantikz}
\end{equation*}
```

where $U^\triangle$ is defined below, and $FT^\dagger$ is defined in Section \@ref(quantum-fourier-transform).
::::

In this circuit, we write $U^\triangle$ to mean the unitary that performs function evaluation as
$$
  \ket{x}\ket{u}
  \longmapsto e^{ix\varphi}\ket{x}\ket{u}
$$
or, in circuit language, the gate that acts on the $i$-th register as a controlled-$U^{2^{n-i}}$ gate:

```{r,engine='tikz',engine.opts=list(template="latex/tikz2pdf.tex")}
\newcommand{\vdotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={yshift=0.3em}]{\vdots}}
\newcommand{\ldotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={xshift=-0.7em}]{\ldots}}
\newcommand{\ddotsgate}{\gate[nwires={1},style={fill=white,draw=white},label style={xshift=-0.7em,yshift=0.3em}]{\ddots}}
\begin{equation*}
  \begin{quantikz}
    & \ctrl{4}
    & \qw
  \\
    & \ctrl{3}
    & \qw
  \\
    & \vdotsgate
  \\
    & \ctrl{1}
    & \qw
  \\[2em]
    & \gate{U^\triangle}
    & \qw
  \end{quantikz}
  \qquad=\qquad
  \begin{quantikz}
    & \ctrl{4}
    & \qw
    & \qw
    & \ldotsgate
    & \qw
    & \qw
  \\
    & \qw
    & \ctrl{3}
    & \qw
    & \ldotsgate
    & \qw
    & \qw
  \\
    & \vdotsgate
    & \vdotsgate
    &
    & \ddotsgate
    &
    &
  \\
    & \qw
    & \qw
    & \qw
    & \ldotsgate
    & \ctrl{1}
    & \qw
  \\[2em]
    & \gate{U^{2^{n-1}}}
    & \gate{U^{2^{n-2}}}
    & \qw
    & \ldotsgate
    & \gate{U}
    & \qw
  \end{quantikz}
\end{equation*}
```

It is important to note, however, that we are sweeping something under the rug here.
If we can treat $U^\triangle$ as an oracle itself, then this does indeed give an efficient solution, but if all we have access to is $U$ then things are definitely not efficient: we are calling it $2^{n-1}+2^{n-2}+\ldots+2^0=2^n-1$ many times, which is *exponentially bad*!





## Quantum Fourier transform {#quantum-fourier-transform}

<div class="video" title="Hadamard and quantum Fourier transforms" data-videoid="qF9Wpgaqqww"></div>

TO-DO: going to define $FT^\dagger$ from \@ref(phase-estimation)

TO-DO: just a small little detour back into more foundational/less applied stuff

TO-DO: BUT we're going to focus on the "circuit story" instead of the mathematical one (i.e. no mention of actual Fourier theory!)

::: {.todo}
<!-- TO-DO: finish -->
:::

::: {.todo}
<!-- .technical make sure to mention that the Hadamard transform is a special case of the Fourier transform over the group $\mathbb{Z}_2^n$ -->
:::





## Hidden order determination

TO-DO: see 7.5 in v5

TO-DO: note that this is useful for Shor's

::: {.todo}
<!-- TO-DO: finish -->
:::






## Shor's algorithm {#shors-algorithm}

<div class="video" title="Shor's algorithm" data-videoid="TioipZX7d5U"></div>

::: {.todo}
<!-- TO-DO -->
:::





## *Remarks and exercises* {#remarks-and-exercises-algorithms}


### RSA

<div class="video" title="RSA" data-videoid="1rtagYeUvFo"></div>

::: {.todo}
<!-- TO-DO -->
:::


### More complexity classes

<div class="video" title="The power of quantum" data-videoid="3Z29RJ40bVw"></div>

::: {.todo}
<!-- TO-DO -->
:::


### Implementing reflections

::: {.todo}
<!-- TO-DO -->
:::


### Grover's optimality

::: {.todo}
<!-- TO-DO: optimality of Grover -->
:::
